---
title: "Reimagining NBA Player Categorizations Through Clustering"
author: "Yashwant Sathish Kumar"
date: "03/03/2024"
output: html_document
---

```{r include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

### Loading Libraries

```{r warning=FALSE} 
# devtools::install_github("abresler/nbastatR")
library("nbastatR")
library(stats)
library(mclust)
library(mdthemes)
library(gghighlight)
library(factoextra)
library(extrafont)
# font_import(prompt=FALSE)
loadfonts() 
library(ggsci) 
library(broom)
library(igraph)
library(tidyverse)
```

```{r}
Sys.setenv("VROOM_CONNECTION_SIZE" = 131072 * 2)
```

### Scraping Data From Basketball Reference Using nbastatR Library
```{r warnings=FALSE, include=FALSE}
# pull seasonal player data 

# advanced stats pulled
advanced_stats <- bref_players_stats(seasons=2016:2024,
                                   tables = c("advanced", "totals"), 
                                   include_all_nba=FALSE,
                                   only_totals=TRUE, nest_data=FALSE,
                                   assign_to_environment=TRUE,
                                   widen_data=TRUE, join_data=TRUE,
                                   return_message=TRUE
                                   ) 

# regular stats pulled
player_stats <- bref_players_stats(seasons=2016:2024,
                                   tables=c('totals', "per_minute"), # totals + per minute stats
                                   include_all_nba=FALSE,
                                   only_totals=TRUE, nest_data=FALSE,
                                   assign_to_environment=TRUE,
                                   widen_data=TRUE, join_data=TRUE,
                                   return_message=TRUE
                                   ) |>
 inner_join(advanced_stats, by= c("idPlayerNBA", "yearSeason"))

## final dataframe containing all data
player_stats <- player_stats |>
    select(namePlayer.x,
         groupPosition.x, yearSeason, slugPosition.x, agePlayer.x, slugTeamBREF.x,
               idPlayerNBA, minutesTotals.x, pctFG.x, pctFG3.x, pctFG2.x, pctEFG.x, pctFT.x,
               fgmPerMinute,  fgaPerMinute, fg3mPerMinute, fg3aPerMinute, fg2mPerMinute,fg2aPerMinute,
               ftmPerMinute, pctFTRate, pctORB, pctDRB, pctTRB, pctAST, pctUSG,
               pctSTL, pctBLK, pctTOV, pfPerMinute, ptsPerMinute) |>
    rename(namePlayer = namePlayer.x, groupPosition = groupPosition.x,
           slugPosition = slugPosition.x, agePlayer = agePlayer.x, slugTeamBREF = slugTeamBREF.x, minutesTotals = minutesTotals.x, pctFG = pctFG.x, pctFG3 = pctFG3.x, pctFG2 = pctFG2.x, pctEFG = pctEFG.x, pctFT = pctFT.x)

```


### Minute Frequency Distribution to Determine Threshold to Filter Player Data
```{r} 
# creating a theme 
theme_kdd <- function() {
  theme(text = element_text(color="blue"),
        axis.title = element_text(face='bold'), # make all axis titles bold
        plot.title = element_text(face='bold', hjust=0.5), # make all plot titles bold
        legend.title = element_text(face='bold'), # make all legend titles bold
        plot.subtitle = element_text(face='italic', hjust=0.5)) # make all subtitles italic
}

# creating frequency histogram
player_stats %>%
  ggplot(aes(x=minutesTotals)) + 
  geom_histogram(aes(y=..density..), position='identity', 
                 fill="#232D4B", alpha=0.9, bins=30) + 
  geom_density(alpha=0.2, fill="#F84C1E", color="#F84C1E") + 
  labs(y = "Frequency", x = "Total Minutes Played", 
       title = "Distribution of Total Minutes Played in a Season") + 
  scale_x_continuous(breaks=seq(-500, 3000, 500)) + 
  theme_minimal() + theme_kdd() + 
  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank(), 
        axis.text.x = element_text(size=10)) 
```

### Data Cleaning
```{r}
# filter data to players that played > 1000 minutes per season
player_stats <- player_stats %>%
  filter(minutesTotals > 1000, idPlayerNBA!=1628977, idPlayerNBA!=1628998) # remove players who's ID repeats

# vector of unique player IDs
unique_playerIDs <- unique(player_stats$idPlayerNBA)
```

### Standardizing Data into Z-Scores Relative to Each Column
```{r}

scaled_stats<- player_stats %>% 
  # select predictor variables defined above
  select(pctFG3, pctFG2, pctFT, 
                fg3aPerMinute, fg2aPerMinute, pctFTRate, 
                pctORB, pctDRB, pctAST, pctUSG,
                pctSTL, pctBLK, pctTOV, 
                ptsPerMinute
                ) %>%
  # scale data into z-scores
  scale()
```

### Determining Number of Clusters For K-Means Using Elbow Method
```{r}
set.seed(222) 
MAX_K <- 20 # max number of clusters
sse <- c() # vector to hold sum of squared errors of each model

for (k in 1:MAX_K) {
  alg <- kmeans(scaled_stats, centers=k, nstart=22, iter.max=20) # running k means algorithm!
  sse <- c(sse, alg$tot.withinss) # get SSE
} 
```

#### Creating Elbow Distribution -- choose 10 distribution
```{r}
tibble(k = 1:MAX_K, SSE = sse) %>%
  ggplot(aes(x=k, y=SSE)) + 
  geom_point(color="#F84C1E") + geom_line(color="#232D4B") +
  labs(x = "K", y = "SSE", title = "Where does this level off?") + # set axis/plot titles
  scale_x_continuous(breaks=seq(1, MAX_K, 1)) + # define x-axis
  theme_minimal() + theme_kdd() + # add themes
  theme(panel.grid.minor.x = element_blank(), panel.grid.minor.y = element_blank()) # manually alter theme
```
#### Choosing 10 clusters
```{r}
set.seed(22)
# re-run K-Means with 10 clusters
K <- 10
kmeans10 <- kmeans(scaled_stats, centers=K, nstart=22, iter.max=20)
km_centers <- as.data.frame(kmeans10$centers) # SCALED cluster centers/means

# name clusters 
cluster_names <- paste0('Cluster ', 1:K)

km_centers$Cluster <- cluster_names

# renaming predictor attributes for interpretability on graph
km_centers <- km_centers %>%
  rename(c('AST'='pctAST', 'BLK'='pctBLK', # give predictors a shorter name for plotting
           'DRB'='pctDRB', '2PA'='fg2aPerMinute',
           '3PA'='fg3aPerMinute', 'FTA'='pctFTRate', 'USG' = 'pctUSG',
           'ORB'='pctORB', 'PTS'='ptsPerMinute', 'STL'='pctSTL',
           'TOV'='pctTOV', 'FT%'='pctFT', '2P%'='pctFG2', '3P%'='pctFG3')) %>% 
  pivot_longer(!Cluster, names_to = 'feature', values_to = 'z_val') # pivot data to make plotting easier

# reset the order of predictor variables for plotting
km_centers$feature <- factor(km_centers$feature, levels=c('PTS', 'AST', 'USG', 'ORB', 'DRB', 
                                                          'STL','BLK', 'TOV', '2PA', 
                                                          '3PA', 'FTA','2P%', '3P%', 
                                                          'FT%')) 

km_centers$Cluster <- factor(km_centers$Cluster, levels=cluster_names)
```

### Plotting Example Cluster (6)
```{r}
km_centers %>% 
  ggplot(aes(x=feature, y=z_val, color=Cluster)) + 
  geom_point(color="black", size=3) + 
  gghighlight(Cluster == 'Cluster 7', use_direct_label = FALSE) + # highlight cluster 6
  labs(x = "Predictor", y = "Cluster Center",  
       title = "Visualizing K-Means Cluster Makeups", 
       subtitle = "Cluster 6: ") + 
  theme_minimal() + theme_kdd() + 
  theme(legend.position = "none", 
        axis.text.x = element_text(angle=45, size=10))

```

### Viewing Players Grouped into a Specific Cluster
```{r}
tibble(cluster=kmeans10$cluster, name=player_stats$namePlayer, season=player_stats$yearSeason) %>%
  filter(cluster == 7)
```


### Visualized all Centers for Each Cluster
```{r warning=FALSE}
km_centers %>% 
  ggplot(aes(x=feature, y=z_val, color=Cluster)) + 
  geom_point() + # plot points
  scale_color_brewer(palette="Paired") + # color points
  gghighlight(use_direct_label = FALSE) + # highlight each cluster
  facet_wrap(~ Cluster, ncol=3) + # create seperate plots for each cluster
  labs(x = "Predictor", y = "Cluster Center", 
       title = "K-Means Cluster Makeups Visual") + 
  theme_minimal() + theme_kdd() + 
  theme(legend.position = "none", strip.text = element_text(face='bold'),
        axis.text.x = element_text(angle = 45, size=8), # alter axis text
        panel.grid.minor = element_blank())
```

Cluster 1: Interior Slashers

- High frequency of 2-Point Shots
- High 2-Point Efficiency
- High frequency of free-throws drawn

Cluster 2: Pass-First Playmakers

- Rarely contribute to scoring
- High Assist (Passing) Rate
- High Turnover Rates

Cluster 3: Rebounding Big Men

- High Rebounding Rates
- High 2-Point %
- High 2-Point Attempts

Cluster 4: Elite Perimeter Scorers
- High scoring, assists, usage 
- High 3-Point Percentage

Cluster 5: Pure Shooters

- High 3-Point Percentage
- High 3-Point Attempts
- Low rankings across other cateogories

Cluster 6: "3 & D" Players

- High 3-Point Percentage
- High 3-Point Attempts
- High Block Rate

Cluster 7: Average Perimeter Scorers

- High 3-Point Percentage
- High 3-Point Attempts
- Low Rankings Across Board for Others

Cluster 8: High-Energy Rebounders
- High Offensive Rebounding Rates

Cluster 9: Defensive Stoppers
- High Steal Rates
- High Rebounding Rates

Cluster 10: Offensive Generators
- High Scoring Rates
- High Assist Rates
- High Usage Rate











































